---
title: "R Notebook"
output: html_notebook
---

## Practice Thinking Questions

1.  The participants are expected to download this data from the provided link, and explore the data set structure, distribution of sampling locations, and answer which area in Bangalore is the most popular for birding?

2.  For a species of your choice in Bangalore create plots that demonstrate the seasonality of a species, or at least list down the steps required to do so.

3.  For a given region find areas which are promising for conservation but do not show up as important on the hotspot map.

4.  See how birder behavior and therefore the data is impacted by events such as COVID?

5.  Create a graph of seasonality of Eastern Cattle Egret in Bangalore.

    ### Import Data

    ```{r}
    library(tidyverse)
    library(sf)
    library(terra)
    library(dplyr)
    library(auk)
    library(lubridate)

    #The pipe operator |> and %>%

    # Raw data files (huge files, don’t open in Excel!)
    ebd_file <- "ebd_IN-KA-BN_202401_202412_smp_relAug-2025/ebd_IN-KA-BN_202401_202412_smp_relAug-2025.txt"
    smp_file <- "ebd_IN-KA-BN_202401_202412_smp_relAug-2025/ebd_IN-KA-BN_202401_202412_smp_relAug-2025_sampling.txt"

    ebd <- auk_ebd("ebd_IN-KA-BN_202401_202412_smp_relAug-2025/ebd_IN-KA-BN_202401_202412_smp_relAug-2025.txt", file_sampling = "ebd_IN-KA-BN_202401_202412_smp_relAug-2025/ebd_IN-KA-BN_202401_202412_smp_relAug-2025_sampling.txt")
    ```

<!-- -->

1.  Explore the data set structure, distribution of sampling locations,

    ```{r}

    # looking at the data
    ebd_data <- read.delim(file = "ebd_IN-KA-BN_202401_202412_smp_relAug-2025/ebd_IN-KA-BN_202401_202412_smp_relAug-2025.txt", sep = "\t")

    ebd_sample <- read.delim(file = "ebd_IN-KA-BN_202401_202412_smp_relAug-2025/ebd_IN-KA-BN_202401_202412_smp_relAug-2025_sampling.txt", sep = "\t")
    ```

    (ii) Spatial distribution of sampling locations

    ```{r}

    library(mapview)
    library(sf)

    ebird_sf <- st_as_sf(
      ebd_sample,
      coords = c("LONGITUDE", "LATITUDE"),
      crs = 4326   # WGS84 lat-long
    )

    # Quick interactive map
    mapview(ebird_sf, 
            cex = 2,             # point size
            alpha = 0.6)  

    ```

2.  Answer which area in Bangalore is the most popular for birding?

For this we have too many points and clutter in the map currently, so lets filter and re-visualise

```{r}

ebd <- auk_ebd("ebd_IN-KA-BN_202401_202412_smp_relAug-2025/ebd_IN-KA-BN_202401_202412_smp_relAug-2025.txt", file_sampling = "ebd_IN-KA-BN_202401_202412_smp_relAug-2025/ebd_IN-KA-BN_202401_202412_smp_relAug-2025_sampling.txt")

ebd_filters <- ebd %>% 
  auk_date(date = c("2020-06-01", "2025-06-30")) %>% 
  # restrict to the standard traveling and stationary count protocols
  auk_protocol(protocol = c("Stationary", "Traveling")) %>% 
  auk_complete()
ebd_filters

#define the output file
f_ebd <- file.path("results/ebd_filtered.txt")
f_sampling <- file.path("results/ebd_filtered_sampling.txt")

# only run if the files don't already exist
if (!file.exists(f_ebd)) {
  auk_filter(ebd_filters, file = f_ebd, file_sampling = f_sampling)
}


#Read into R as a dataframe
ebd_filt_sam <- read.delim("results/ebd_filtered_sampling.txt", sep = "\t")
# Convert to sf object
ebd_sam_sf <- st_as_sf(
  ebd_filt_sam,
  coords = c("LONGITUDE", "LATITUDE"), # column names in your ebd_df
  crs = 4326,                           # WGS84
  remove = FALSE                        # keep lat/long columns
)

# Visualize with mapview again
mapview(ebd_sam_sf, 
        cex = 2,             # point size
        alpha = 0.6) 
st_write(ebd_sam_sf, dsn = "ebd_sam_sf.kml", driver = "KML")
```

Now we will use a grid based approach to understand the distribution

```{r}
# Reproject to India Albers Equal Area 
ebd_sf_proj <- st_transform(ebd_sam_sf, 7755)   # EPSG:7755
#also do it for sampling data

# Make grid with 1 km cells
grid <- st_make_grid(ebd_sf_proj, cellsize = 1000, square = TRUE)

# Convert grid to sf
grid_sf <- st_sf(grid_id = 1:length(grid), geometry = grid)

# Spatial join: assign points to grid cells
joined <- st_join(ebd_sf_proj, grid_sf, join = st_within)

# Count per grid cell
grid_counts <- joined %>%
  st_drop_geometry() %>%
  count(grid_id, name = "checklist_count")

# Add counts back to grid
grid_sf <- left_join(grid_sf, grid_counts, by = "grid_id")

# Replace NA with 0
grid_sf$checklist_count[is.na(grid_sf$checklist_count)] <- 0

library(ggplot2)

ggplot(grid_sf) +
  geom_sf(aes(fill = checklist_count), color = "grey30", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", trans = "sqrt") +
  theme_minimal() +
  labs(fill = "# Checklists", title = "eBird Checklists per Grid Cell")
mapview(grid_sf, zcol = "checklist_count", legend = TRUE)

st_write(grid_sf, "results/grid_counts.shp")

# Now we visualize
```

**Now we will try to figure out the distribution of biodiversity**

```{r}
#read ebd
ebd <- read_tsv("results/ebd_filtered.txt")
# convert to sf and reproject 

ebd_sf <- st_as_sf(
  ebd,
  coords = c("LONGITUDE", "LATITUDE"),
  crs = 4326)

ebd_sf_proj <- st_transform(ebd_sf, 7755)   # India Albers

#Join points to grids
joined <- st_join(ebd_sf_proj, grid_sf, join = st_within)

## compute checklist richness per grid
# Each checklist_id has multiple species entries, so count unique species
checklist_richness <- joined %>%
  st_drop_geometry() %>%
  group_by(grid_id, `SAMPLING.EVENT.IDENTIFIER`) %>%
  summarise(species_richness = n_distinct(`SCIENTIFIC.NAME`), .groups = "drop")

# get the maximum richness per grid
grid_richness <- checklist_richness %>%
  group_by(grid_id) %>%
  summarise(max_species = max(species_richness, na.rm = TRUE), .groups = "drop")

# Add back to grid
grid_sf <- left_join(grid_sf, grid_richness, by = "grid_id")

# Replace NA with 0
grid_sf$max_species[is.na(grid_sf$max_species)] <- 0

## Plot the maximum species richness

ggplot(grid_sf) +
  geom_sf(aes(fill = max_species), color = "grey30", size = 0.1) +
  scale_fill_viridis_c(option = "magma", trans = "sqrt") +
  theme_minimal() +
  labs(fill = "Max species per checklist",
       title = "Maximum Species Reported per Grid Cell")
mapview(grid_sf, zcol = "max_species", legend = TRUE)
st_write(grid_sf, "results/grid_species_counts.shp")

```

```         
```

**Optional exercise:** For a species of your choice in Bangalore create plots that demonstrate the seasonality of a species, or at least list down the steps required to do so.

Before we jump into seasonality we will perform spatial thinning

Lets choose the species given - western

```{r}

install.packages("spThin")
library(spThin)
#extract localities
locs <- select(ebd_filt_sam, LATITUDE, LONGITUDE, SAMPLING.EVENT.IDENTIFIER)
thin(
  loc.data = locs,
  lat.col = "LATITUDE",
  long.col = "LONGITUDE",
  spec.col = "SAMPLING.EVENT.IDENTIFIER",
  thin.par = 0.2,           # thinning distance in km
  reps = 1,              # number of replicates
  locs.thinned.list.return = TRUE,
  write.files = FALSE,     # don’t save files automatically
  verbose = TRUE
)


thinned <- thin(
  loc.data = ebird_data,
  lat.col = "latitude",
  long.col = "longitude",
  spec.col = "species",
  thin.par = 0.2,
  reps = 1,
  locs.thinned.list.return = TRUE,
  write.files = FALSE
)

#### alternate way ####

# Set grid size
gridsize <- 200  # in meters

# 2. Create 200 m grid across the data range
grid <- st_make_grid(ebd_sf_proj, cellsize = gridsize, what = "polygons")
grid_sf <- st_sf(grid_id = 1:length(grid), geometry = grid)

# 5. Spatial join: assign each checklist to a grid cell
ebird_joined <- st_join(ebd_sf_proj, grid_sf, join = st_intersects)

# 6. Create unique site IDs using coordinate pairs
ebird_joined$site_id <- paste0(round(st_coordinates(ebird_joined)[,1], 2), "_",
                               round(st_coordinates(ebird_joined)[,2], 2))
# 7. Count visits per site within each grid cell
most_visited <- ebird_joined %>%
    count(grid_id, site_id) %>%
    group_by(grid_id) %>%
    slice_max(n, with_ties = FALSE)
# 8 Filter original data to retain only those sites
background_bias <- ebird_joined %>%
    filter(site_id %in% most_visited$site_id & grid_id %in% most_visited$grid_id) %>%
    group_by(grid_id) %>%
    slice(1) %>%
    ungroup()

# 9. Save to file (optional)
st_write(background_bias, "thinned sampling.gpkg")
background_bias_coords <- background_bias %>%
    mutate(
        longitude = st_coordinates(.)[, 1],
        latitude = st_coordinates(.)[, 2]
    ) %>%
    st_drop_geometry()

# Save as CSV
write_tsv(background_bias, "results/thinnged_background_200m.txt")


```

3.  **Now generating seasonality of the species**

    ```{r}
    ### generating detection frequencies ###
    #First we zerofill
    ebd_filtered <- read_tsv("results/ebd_filtered.txt")
    ebd_egret <- ebd_filtered %>% filter(`SCIENTIFIC NAME` == "Ardea coromanda")
    write_tsv(ebd_egret, "results/ebd_egret.txt")
    ebd_smp <- read_sampling("results/ebd_filtered_sampling.txt")
    ebd_egret <- read_ebd("results/ebd_egret.txt")
    ebd_egret <- filter(ebd_egret, checklist_id %in% ebd_smp$checklist_id)
    ebd_zf <- auk_zerofill(ebd_egret, ebd_smp, collapse = TRUE)

    write_tsv(ebd_zf, "results/egret_zf.txt")
    #filter the species
    # now split the date format into month and year
    ebd_zf <- ebd_zf %>%
      mutate(
        year = year(observation_date),
        month = month(observation_date),
        day = day(observation_date)
      )
    # Group by month and species, calculate detections and total checklists
    monthly_freq <- ebd_zf %>%
        group_by(month) %>%
        summarize(
            total_detections = sum(species_observed),
            total_checklists = n(),
            frequency = total_detections / total_checklists
        ) %>%
        ungroup()
    monthly_freq$perct <- monthly_freq$frequency*100
    monthly_freq <- na.omit(monthly_freq)    
        
    # Plot the frequency of detections for each species
    p<- ggplot(monthly_freq, aes(x = month, y = perct)) +
        geom_line() +
        geom_point() +
        scale_x_continuous(breaks = seq(1, 12, by = 1), limits = c(1, 12)) +
        labs(
            title = "Frequency of Detections in %",
            x = "Month",
            y = "Frequency of Detections",
            color = "Species"
        )
    plot(p)
    ggsave(p, filename = paste0("cuce-detection", ".png"),
           width = 8, height = 5, units = "in", dpi = 300)


    ```

We will finally, plot the presence absence distribution of the eastern cattle egret

```{r}
zf_egret_df <- st_as_sf(ebd_zf, coords = c("longitude", "latitude"), crs = 4326)

mapview(zf_egret_df, zcol = "species_observed", col.regions = c("grey", "blue"), cex = 2)

state <- st_read("karnataka.shp")

ggplot() +
  geom_sf(data = state, fill = "white", color = "black") +
  geom_sf(data = zf_egret_df, aes(color = species_observed), size = 2) +
  scale_color_manual(values = c("FALSE" = "grey", "TRUE" = "red"),
                     labels = c("Absence", "Presence"),
                     name = "Observation") +
  theme_minimal() +
  labs(title = "Presence–Absence of Eastern Cattle-egret in Bengaluru")


```
