---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

## Practice Thinking Questions

1.  The participants are expected to download this data from the
    provided link, and explore the data set structure, distribution of
    sampling locations, and answer which area in Bangalore is the most
    popular for birding?

2.  For a species of your choice in Bangalore create plots that
    demonstrate the seasonality of a species, or at least list down the
    steps required to do so.

3.  For a given region find areas which are promising for conservation
    but do not show up as important on the hotspot map.

4.  See how birder behavior and therefore the data is impacted by events
    such as COVID?

5.  Create a graph of seasonality of Eastern Cattle Egret in Bangalore.

    ```{r}

    library(sf) 
    library(terra) 
    library(dplyr) 
    library(auk) 
    library(lubridate)

    #Raw data files (huge files, don’t open in Excel!)

    ebd_file <- "ebd_IN-KA-BN_202401_202412_smp_relAug-2025/ebd_IN-KA-BN_202401_202412_smp_relAug-2025.txt" 
    smp_file <- "ebd_IN-KA-BN_202401_202412_smp_relAug-2025/ebd_IN-KA-BN_202401_202412_smp_relAug-2025_sampling.txt"

    ebd <- auk_ebd("ebd_IN-KA-BN_202401_202412_smp_relAug-2025/ebd_IN-KA-BN_202401_202412_smp_relAug-2025.txt", file_sampling = "ebd_IN-KA-BN_202401_202412_smp_relAug-2025/ebd_IN-KA-BN_202401_202412_smp_relAug-2025_sampling.txt")
    ebd
    ```

6.  Explore the data set structure, distribution of sampling locations,

    ```{r}

    # looking at the data
    ebd_data <- read.delim(file = "ebd_IN-KA-BN_202401_202412_smp_relAug-2025/ebd_IN-KA-BN_202401_202412_smp_relAug-2025.txt", sep = "\t")

    ebd_sample <- read.delim(file = "ebd_IN-KA-BN_202401_202412_smp_relAug-2025/ebd_IN-KA-BN_202401_202412_smp_relAug-2025_sampling.txt", sep = "\t")
    ```

    (ii) Spatial distribution of sampling locations

    ```{r}

    library(mapview)
    library(sf)

    ebird_sf <- st_as_sf(
      ebd_sample,
      coords = c("LONGITUDE", "LATITUDE"),
      crs = 4326   # WGS84 lat-long
    )

    # Quick interactive map
    mapview(ebird_sf, 
            cex = 2,             # point size
            alpha = 0.6,
            map.types = "OpenStreetMap")  

    ```

7.  Answer which area in Bangalore is the most popular for birding?

For this we have too many points and clutter in the map currently, so
lets filter and re-visualise

```{r}
#set ebd path

library(data.table)
library(dplyr)

# define paths
ebd_path <- "ebd_IN-KA-BN_202401_202412_smp_relAug-2025/ebd_IN-KA-BN_202401_202412_smp_relAug-2025.txt"
sampling_path <- "ebd_IN-KA-BN_202401_202412_smp_relAug-2025/ebd_IN-KA-BN_202401_202412_smp_relAug-2025_sampling.txt"

f_ebd <- file.path("results/ebd_filtered.txt")
f_sampling <- file.path("results/ebd_filtered_sampling.txt")

dir.create("results", showWarnings = FALSE)

# only run if the filtered files don't already exist
if (!file.exists(f_ebd)) {
  
  # read in the full EBD file (tab-delimited)
  ebd <- fread(ebd_path, sep = "\t", quote = "", na.strings = "")
  
  # apply the same filters as auk_date(), auk_protocol(), auk_duration(), auk_complete()
  ebd_filtered <- ebd %>%
    mutate(`OBSERVATION DATE` = as.Date(`OBSERVATION DATE`)) %>%
    filter(
      `OBSERVATION DATE` >= as.Date("2020-06-01"),
      `OBSERVATION DATE` <= as.Date("2025-06-30"),
      `PROTOCOL TYPE` %in% c("Stationary", "Traveling"),
      `DURATION MINUTES` <= 60,
      `ALL SPECIES REPORTED` == 1
    )
  
  # write out filtered EBD file
  fwrite(ebd_filtered, f_ebd, sep = "\t", na = "", quote = FALSE)
  
  # do the same filtering on the sampling file
  sampling <- fread(sampling_path, sep = "\t", quote = "", na.strings = "")
  
  sampling_filtered <- sampling %>%
    mutate(`OBSERVATION DATE` = as.Date(`OBSERVATION DATE`)) %>%
    filter(
      `OBSERVATION DATE` >= as.Date("2020-06-01"),
      `OBSERVATION DATE` <= as.Date("2025-06-30"),
      `PROTOCOL TYPE` %in% c("Stationary", "Traveling"),
      `DURATION MINUTES` <= 60,
      `ALL SPECIES REPORTED` == 1
    )
  
  fwrite(sampling_filtered, f_sampling, sep = "\t", na = "", quote = FALSE)
}
#Read into R as a dataframe
ebd_filt_sam <- read.delim("results/ebd_filtered_sampling.txt", sep = "\t")
# Convert to sf object
ebd_sam_sf <- st_as_sf(
  ebd_filt_sam,
  coords = c("LONGITUDE", "LATITUDE"), # column names in your ebd_df
  crs = 4326,                           # WGS84
  remove = FALSE                        # keep lat/long columns
)

# Visualize with mapview again
mapview(ebd_sam_sf, 
        cex = 2,             # point size
        alpha = 0.6,
        map.types = "OpenStreetMap") 

```

Now we will use a grid based approach to understand the distribution

```{r}
#st_write(ebd_sam_sf, dsn = "ebd_sam_sf.kml", driver = "KML")
# Reproject to India Albers Equal Area 
ebd_sf_proj <- st_transform(ebd_sam_sf, 7755)   # EPSG:7755
#also do it for sampling data

# Make grid with 1 km cells
grid <- st_make_grid(ebd_sf_proj, cellsize = 1000, square = TRUE)

# Convert grid to sf
grid_sf <- st_sf(grid_id = 1:length(grid), geometry = grid)

# Spatial join: assign points to grid cells
joined <- st_join(ebd_sf_proj, grid_sf, join = st_within)

# Count per grid cell
grid_counts <- joined %>%
  st_drop_geometry() %>%
  count(grid_id, name = "checklist_count")

# Add counts back to grid
grid_sf <- left_join(grid_sf, grid_counts, by = "grid_id")

# Replace NA with 0
grid_sf$checklist_count[is.na(grid_sf$checklist_count)] <- 0

library(ggplot2)

ggplot(grid_sf) +
  geom_sf(aes(fill = checklist_count), color = "grey30", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", trans = "sqrt") +
  theme_minimal() +
  labs(fill = "# Checklists", title = "eBird Checklists per Grid Cell")


# Now we visualize
```

```{r}
#st_write(grid_sf, "results/grid_counts.shp")

mapview(grid_sf, zcol = "checklist_count", legend = TRUE,  map.types = "OpenStreetMap")
```

**Now we will try to figure out the distribution of biodiversity**

```{r}

library(readr)

#read ebd
ebd <- read_tsv("results/ebd_filtered.txt")
# convert to sf and reproject 

ebd_sf <- st_as_sf(
  ebd,
  coords = c("LONGITUDE", "LATITUDE"),
  crs = 4326)

ebd_sf_proj <- st_transform(ebd_sf, 7755)   # India Albers

#Join points to grids
joined <- st_join(ebd_sf_proj, grid_sf, join = st_within)

## compute checklist richness per grid
# Each checklist_id has multiple species entries, so count unique species
checklist_richness <- joined %>%
  st_drop_geometry() %>%
  group_by(grid_id, `SAMPLING EVENT IDENTIFIER`) %>%
  summarise(species_richness = n_distinct(`SCIENTIFIC NAME`), .groups = "drop")

# get the maximum richness per grid
grid_richness <- checklist_richness %>%
  group_by(grid_id) %>%
  summarise(max_species = max(species_richness, na.rm = TRUE), .groups = "drop")

# Add back to grid
grid_sf <- left_join(grid_sf, grid_richness, by = "grid_id")

# Replace NA with 0
grid_sf$max_species[is.na(grid_sf$max_species)] <- 0

## Plot the maximum species richness

ggplot(grid_sf) +
  geom_sf(aes(fill = max_species), color = "grey30", size = 0.1) +
  scale_fill_viridis_c(option = "magma", trans = "sqrt") +
  theme_minimal() +
  labs(fill = "Max species per checklist",
       title = "Maximum Species Reported per Grid Cell")
```

```{r}
#st_write(grid_sf, "results/grid_species_counts.shp")
mapview(grid_sf, zcol = "max_species", legend = TRUE, map.types = "OpenStreetMap")

```

**Optional exercise:** For a species of your choice in Bangalore create
plots that demonstrate the seasonality of a species, or at least list
down the steps required to do so.

Before we jump into seasonality we will perform spatial thinning

Lets choose the species given - western

```         
```

3.  **Now generating seasonality of the species**

    ```{r}
    ### generating detection frequencies ###
    #First we zerofill
    ebd_filtered <- read_tsv("results/ebd_filtered.txt")
    ebd_egret <- ebd_filtered %>% filter(`SCIENTIFIC NAME` == "Ardea coromanda")
    #write_tsv(ebd_egret, "results/ebd_egret.txt")
    ebd_smp <- read_sampling("results/ebd_filtered_sampling.txt")
    ebd_egret <- read_ebd("results/ebd_egret.txt")
    ebd_egret <- filter(ebd_egret, checklist_id %in% ebd_smp$checklist_id)
    ebd_zf <- auk_zerofill(ebd_egret, ebd_smp, collapse = TRUE)

    #write_tsv(ebd_zf, "results/egret_zf.txt")
    #filter the species
    # now split the date format into month and year
    ebd_zf <- ebd_zf %>%
      mutate(
        year = year(observation_date),
        month = month(observation_date),
        day = day(observation_date)
      )
    # Group by month and species, calculate detections and total checklists
    monthly_freq <- ebd_zf %>%
        group_by(month) %>%
        summarize(
            total_detections = sum(species_observed),
            total_checklists = n(),
            frequency = total_detections / total_checklists
        ) %>%
        ungroup()
    monthly_freq$perct <- monthly_freq$frequency*100
    monthly_freq <- na.omit(monthly_freq)    
        
    # Plot the frequency of detections for each species
    p<- ggplot(monthly_freq, aes(x = month, y = perct)) +
        geom_line() +
        geom_point() +
        scale_x_continuous(breaks = seq(1, 12, by = 1), limits = c(1, 12)) +
        labs(
            title = "Frequency of Detections in %",
            x = "Month",
            y = "Frequency of Detections",
            color = "Species"
        )
    plot(p)


    ```

```{r}
ggsave(p, filename = paste0("egret season", ".png"),
       width = 8, height = 5, units = "in", dpi = 300)

zf_egret_df <- st_as_sf(ebd_zf, coords = c("longitude", "latitude"), crs = 4326)

mapview(zf_egret_df, zcol = "species_observed", col.regions = c("grey", "blue"), cex = 2,  map.types = "OpenStreetMap")



```

```{r}

```
